---
title: "Chapter 1 - Matrix Algebra"
format: html
---

```{r setup}
#| echo: false
#| message: false
library(matlib)
```

## Basic Definitions


## Properties of Matrices


## Linear independence


## Rank


## Orthogonal Matrices



## Exercises

For exercises 1-5, let

$$
X = \left[ {\begin{array}{cc}
    2 & 3\\
    -1 & 4\\
    \end{array}} \right ];\quad
Y = \left[ {\begin{array}{ccc}
    2 & 0 & 1\\
    1 & -2 & 3\\
    \end{array}} \right ];\quad
Z = \left[ {\begin{array}{cc}
    1 & 1\\
    -1 & 1\\
    0 & 2 \\
    \end{array}} \right ]
$$

$$
W = \left [{\begin{array}{cc}
    1 & 0 \\
    8 & 3 \\
\end{array}}\right];\quad
T = \left [{\begin{array}{cc}
    1 & 0 \\
    0 & 1 \\
\end{array}} \right]
$$

### Problem 1
If possible, perform each matrix operation. If the indicated operation is not possible, explain why.

a.) $X + Y$ - This operation is not possible because $X$ and $Y$ do not have the same dimensions.
b.) $X + W$
```{r}
X <- matrix(c(2,3,
              -1,4), ncol = 2, nrow = 2, byrow = TRUE)
W <- matrix(c(1,0,
              8,3), ncol = 2, nrow = 2, byrow = TRUE)

X + W

```
c.) $X - T$
```{r}
T <- matrix(c(1,0,
              0,1), ncol = 2, nrow = 2, byrow = TRUE)

```
$$
X - T = \left[
 \begin{array}{rr}
  1 & 3 \\ 
  -1 & 3\ \\ 
  \end{array}
\right]
$$
d.) $XY$

```{r}
Y <- matrix(c(2,0,1,
              1,-2,3), 
            ncol = 3, nrow = 2, byrow = TRUE)

# printMatrix(X%*%Y, latex = TRUE, tol = 8)
```

$$
\left[
 \begin{array}{rrr}
  7 & -6 & 11 \\ 
  2 & -8 & 11 \\ 
  \end{array}
\right]
$$
e.) $YX$ - This operation is not allowed as $Y$ is a $2\times 3$ matrix and $X$ is $2\times2$. The matrices are not conformable in this order.

f.) $3X$ - is simple scalar multiplication.

$$
3X = \left[ {\begin{array}{cc}
    3*2 & 3*3\\
    3*(-1) & 3*4\\
    \end{array}} \right ] = 
    \left[ {\begin{array}{cc}
    6 & 9\\
    -3 & 12\\
    \end{array}} \right ]
$$

g.) $XT$ - $T$ is just the $2\times2$ identity matrix so $XT = X$.

h.) $TX$ - Again, $T$ is just the identity matrix and can be applied to either "side" of $X$ in a matrix multiplication operation to yield $TX = X$.

i.) $X + (Y+Z)$ - None of these matrices have the same dimensions so this operation is not defined.

j.) $Z(T + W)$ - This operation is defined.

```{r}
Z <- matrix(c(1,1,
              -1,1,
              0,2), ncol = 2, nrow = 3, byrow = TRUE)

# printMatrix(Z %*% (T + W), latex = TRUE, tol = 7)
```
$$
\left[
 \begin{array}{rr}
  10 & 4 \\ 
  6 & 4 \\ 
  16 & 8 \\ 
  \end{array}
\right]
$$
k.) $Y(T + W)$ - This operation is not defined because $Y$ does not conform with $T+W$.

### Problem 2
Find $(W + T)Y$, $WY$, and $TY$. Show that $(W+T)Y = WY + TY$, thus illustrating property 5 for matrix operations.

```{r}
(W + T) %*% Y
```

```{r}
W %*% Y
```

```{r}
T%*%Y
```
From these results, it is clear that $(W+T)Y = WY + TY$.

### Problem 3
Find the zero matrix associated with $X$; of $Y$. Can you generalize to describe the zero matrix for any $n\times k$ matrix?

The zero matrix is defined such that $X + 0 = 0 + X = X$. The matrix that satisfies this condition for $X$ is the $2\times2$ $0$ matrix:
$$
\left[{
\begin{array}{cc}
0 & 0 \\
0 & 0 \\
\end{array}
} \right]
$$
For $Y$, the appropriate zero matrix is
$$
\left[{
\begin{array}{ccc}
0 & 0 & 0\\
0 & 0 & 0\\
\end{array}
} \right]
$$
In general, the zero matrix for any $n\times k$ matrix will be the matrix of the same dimensions with zeroes for all entries.

### Problem 4
Find the negative of $X$; of $Y$. Can you generalise to describe the form of the negative matrix for any $n\times k$ matrix?

The negative matrix, $N$ of a matrix $X$ satisfies the condition that $X + N = 0$. We can obtain this matrix by scalar multiplication by $-1$.

$X + (-1)X = 0$

$$
(-1)X = \left[
 \begin{array}{rr}
  -2 & -3 \\ 
  1 & -4 \\ 
  \end{array}
\right]
$$

```{r}
printMatrix(-1*Y, latex = TRUE, tol=7)
```
$$
(-1)Y = 
\left[
 \begin{array}{rrr}
  -2 & 0 & -1 \\ 
  -1 & 2 & -3 \\ 
  \end{array}
\right]
$$
In general, the negative of a matrix, $A$, is $(-1)A$.

### Problem 5
Find $2Y$, $X(2Y)$, $XY$, and then $2(XY)$. Show that $X(2Y)=2(XY)$, thus illustrating property 6 for matrix operations.

```{r}
printMatrix(X %*% (2*Y), latex = TRUE, tol=7)
```

$$
2Y = \left[
 \begin{array}{rrr}
  4 & 0 & 2 \\ 
  2 & -4 & 6 \\ 
  \end{array}
\right]
$$

$$
X(2Y) = \left[
 \begin{array}{rrr}
  14 & -12 & 22 \\ 
  4 & -16 & 22 \\ 
  \end{array}
\right]
$$

$$
XY = \left[
 \begin{array}{rrr}
  7 & -6 & 11 \\ 
  2 & -8 & 11 \\ 
  \end{array}
\right]
$$

$$
2(XY) = \left[
 \begin{array}{rrr}
  14 & -12 & 22 \\ 
  4 & -16 & 22 \\ 
  \end{array}
\right]
$$
